---
title: "Lighting Safety Analysis"
knit: (function(input_file, encoding) {
  rmarkdown::render(input_file, 
                    encoding = encoding,
                    output_file = 'docs/index.html') })
author: "Mark Egge"
date: "5/3/2020"
output: html_document
---

# Pennsylvania Road Lighting and Safety

This repository demonstrates an end-to-end analysis of the question:

_How does roadway lighting affect roadway safety?_

This project demonstrates the use of the R statistical programming languaeg and several packages to conduct a "data science" analysis of the relationship between roadway lighting (i.e. streetlights) and crashes on major roads in Pennsylvania.

## Concepts Demonstrated

This package illustrates several common transportation data analysis tasks, including:

* Managing data using the fast `data.table` package
* Using the `sf` package to perform a spatial join between point data (crashes) and line data (road segments)
* Using the `rpart` package to predict lighting conditions for road segments with unknown lighting conditions based on road segments with known lighting conditions
* Exploratory data analysis using the `leaflet` mapping package and `ggplot2` plotting package
* Fitting a zero-inflated Poisson regression model to predict crash counts controlling for traffic volume

# Method

Load required packages:

```{r, message=FALSE}
library(sf) # spatial data manipulation
library(data.table) # fast data manipulation
library(leaflet) # easy mapping in R
library(ggplot2) # plotting
library(rpart); library(rpart.plot) # decision tree models
library(MASS); library(pscl) # Poisson and zero-inflated regression

source("R/summary_plots.R") # function for creating summary plots

options(scipen = 99) # disable scientific notation
```

### Load Data

**Road Segments and Traffic:** A shapefile of Pennsylvania traffic data has been prepared from PennDOT's [RMSTRAFFIC](https://data-pennshare.opendata.arcgis.com/datasets/rmstraffic-traffic-volumes) data, retaining only primary state routes. The shapefile is read into R as a spatial object using `st_read` function from the `sf` package:

```{r}
segments <- st_read("shp/traffic/traffic.shp")
segments$id <- 1:nrow(segments) # assign a unique identifier for each segment
```

**Crashes**: Having downloaded the [PA Crash data files for 2012 – 2016](https://crashinfo.penndot.gov/PCIT/welcome.html) (five years is the typical period of analysis for safety investigations), the CSV files are loaded into R using the `fread` function from the `data.table` package. Compared with `read.csv`, `fread` is much faster, and has sensible defaults (e.g. `stringsAsFactors = FALSE`). A list object with each year's data is populated, then the `rbindlist` from `data.table` binds together the rows from each object in the list.

```{r}
crashes <- list()
for(year in as.character(2012:2016)) {
  crashes[[year]] <- fread(paste0("data/Statewide_", year, "/CRASH_", year, "_Statewide.csv"))
}
crashes <- rbindlist(crashes) # join the five files into one
```

The result is a dataset called `crashes` with `r nrow(crashes)` observations.

We then run the provided `summary_plots` function to generate a PDF with a descriptive plot of each attribute in the dataset. 

```{r, echo=FALSE, warning=FALSE}
summary_plots(crashes, "out/plots.pdf")
```

Take a moment to inspect the resulting file [out/plots.pdf](plots.pdf):

* What trends do you see in terms of the seasonality of crashes?
* What data cleaning step would you need to take if you were analyzing crashes by Hour of Day?

Finally, filter our crashes to those occurring at night, and create a "crash_lighting" attribute, based on the metadata in [docs/PA_Crashes_Data_Dictionary.pdf](PA_Crashes_Data_Dictionary.pdf):

```{r}
# Filter to only crashes with "2 - Dark - No Street Lights" or "3 - Dark - Street Lights"
crashes <- crashes[ILLUMINATION %in% c(2, 3)]
crashes[ILLUMINATION == 2, crash_lighting := "unlit"]
crashes[ILLUMINATION == 3, crash_lighting := "lit"]
```

### Spatially join crashes to road segments

Next, we **spatially join crashes to road segments** by buffering the road segments by 50 feet and then performing a spatial join.

Note, for a spatial join to execute property, both the source and destination layers must be in the same *spatial projection*. In this case, we use the *EPSG:2272 NAD83 / Pennsylvania South (ft)* projection, which measures distances in feet.

The `st_as_sf` function from the `sf` package creates a spatial object from the crashes dataset based on the latitude and longitude values in the "DEC_LAT" and "DEC_LONG" fields.

```{r}
crashes <- crashes[!(is.na(DEC_LONG) | is.na(DEC_LAT))] # keep only records with x, y values
crashes <- st_as_sf(crashes, coords = c("DEC_LONG", "DEC_LAT"), crs = 4326) # create sf spatial object
crashes <- st_transform(crashes, 2272) # reproject data to EPSG:2272

# buffer road segments by 50 feet
segment_buffer <- st_buffer(segments, dist = 50, endCapStyle = "FLAT")

# spatially join crashes to road segment data to crashes
joined <- st_join(segment_buffer, crashes, left = FALSE) # inner join
```

Finally, we create a new data.table of the aggregate crash counts by road segment, and merge this new attribute back to our segments data:

```{r}
segment_crashes <- as.data.table(st_drop_geometry(joined)) # convert sf to data.table

crash_counts <- segment_crashes[, .(annual_crashes = .N / 5), by = id] # count annual crashes by segment

# merge crash counts back to segments
segments <- merge(segments, crash_counts, by = "id", all.x = TRUE)

# calculate crash rates - crashes per million VMT
segments[is.na(segments$annual_crashes), ]$annual_crashes <- 0
segments$mvmt <- with(segments, (DLY_VMT * 365) / 1000000) # million annual vehicle miles travelled
segments$rate <- with(segments, annual_crashes / mvmt) # annual crashes per 1m annual vmt
```

### Segment Illumination

Our roadway dataset does not include information about overhead lighting. The section below uses crash data to impute a "lighting" attribute for our road segments, and then uses a *decision tree classifier* to assign lighting to unknown segments based on known segments.

```{r}
# count total crashes by segment id and illumination condition
counts <- segment_crashes[, .(crash_count = .N), by = .(id, crash_lighting)]

segment_counts <- dcast(counts, id ~ crash_lighting,  # long to wide transform
                      value.var = "crash_count", fill = 0)         # filling missing values with zero

# assign segment_lighting based on majority of crashes (lit or unlit)
segment_counts[, segment_lighting := ifelse(lit >= unlit, "lit", "unlit")] 

# join the imputed roadway lighting condition to all road segments
light <- merge(as.data.table(st_drop_geometry(segments)), # create data.table from segments attributes
               segment_counts[, .(id, segment_lighting)], # join result, keeping only segment_lighting field
               by = "id", all.x = TRUE) # join by id, keeping all segments

# Table of "lit", "unlit", and "unknown" lighting_conditon attributes
table(light$segment_lighting, useNA = "ifany")
```

For segments with unknown lighting_condition, predict which road segments are illuminated using an rpart decision tree:

```{r}
fit <- rpart(segment_lighting ~ ST_RT_NO + CTY_CODE + DISTRICT_N + SEG_LNGTH_ + CUR_AADT + DLY_VMT, data = light)

# create a confusion matrix to evaluate classifier performance
actual <- light[!is.na(segment_lighting), segment_lighting]
predicted <- predict(fit, type = "class")
(cm <- as.matrix(table(Actual = actual, Predicted = predicted))) # create the confusion matrix
accuracy <- sum(diag(cm)) / sum(cm) # number of correctly classified instances per class  / number of instances
p <- rowSums(cm) / sum(cm); q <- colSums(cm) / sum(cm)
kappa <- (accuracy - sum(p * q)) / (1 - sum(p * q)) # kappa score indicates moderate agreement
cat("Model is", round(accuracy * 100), "% accurate with a Kappa score of", round(kappa, 2))
```

Our decision tree is reasonably accurate (~75%) and the Kappa score (0.55) indicates moderate evidence that our model outperforms a random chance. We can use this model to assign lighting_condition to the unknown segments.

```{r}
# use decision tree to fill in missing lighting values
predictions <- predict(fit, type = "class", newdata = light[is.na(segment_lighting)])
light[is.na(segment_lighting)]$segment_lighting <- predictions

# Updated table of "lit", "unlit", and "unknown" lighting_conditon attributes
table(light$segment_lighting, useNA = "ifany")

# join lighting attribute back to road segments
segments <- merge(segments, light[, .(id, lighting = segment_lighting)], by = "id", all.x = TRUE)
```


## Exploratory Data Analysis

Exploratory data analysis allows the analyst to become more familiar with the data and informs the selection of subsequent modelling or data visualization steps.

```{r}
# convert spatial object to data.table for easy data manipulation
DT <- as.data.table(st_drop_geometry(segments))
DT <- DT[!is.na(rate)] # several segments have no geom or rate

# plot crash counts vs. vmt
# relationship between crashes and vmt seems fairly linear
ggplot(DT, aes(x = mvmt, y = annual_crashes, color = lighting)) +
  geom_point(alpha = 0.65, size = 1) + ggtitle("Crash Count vs. VMT")
  
```

Visualize the illumination conditions and crash rates:

```{r}
# limit map data to Pittsburgh region and reproject to web mercator for leaflet
leaflet_segments <- segments[which(segments$DISTRICT_N %in% c("11", "12") & !st_is_empty(segments)), ]

fpal <- colorFactor(c("yellow", "brown", "orange"), leaflet_segments$lighting)
qpal <- colorQuantile(palette = "YlOrRd", domain = leaflet_segments$rate, n = 6)

leaflet(data = st_transform(leaflet_segments, 4326)) %>% 
  addProviderTiles(providers$CartoDB.DarkMatter) %>% 
  addPolylines(color = ~fpal(lighting), group = "Lighting") %>%
  addPolylines(color = ~qpal(rate), group = "Crash Rates") %>%  
  addLayersControl(
    overlayGroups = c("Lighting", "Crash Rates"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% hideGroup("Crash Rates") %>%
  addLegend(pal = fpal, values = ~lighting, title = "Lighting") %>%
  addLegend(pal = qpal, values = ~rate, title = "Crash Rate Quantile")
```


```{r}

```

Decision trees are useful tools to determine which dataset attributes our predictive of our outcome variables of interest

```{r}
# What explains / predicts crash counts? A decision tree:
fit <- rpart(annual_crashes ~ DISTRICT_N + SEG_LNGTH_ + CUR_AADT + DLY_VMT,
             data = DT)
rpart.plot(fit)

# What explains / predicts crash rates?
fit <- rpart(rate ~ DISTRICT_N + SEG_LNGTH_ + CUR_AADT + DLY_VMT, data = DT)
rpart.plot(fit)
```

```{r}
# box plot of crash rates
boxplot(rate ~ lighting, data = DT[rate < 2])

# are lighting and AADT related?
cor(DT$DLY_VMT, DT$lighting == "lit") # Yes - weak negative correlation

# some descriptive plots
hist(DT$rate, main = "Distribution of Crash Rates")
hist(log(DT$rate), main = "Distribution of Log Crash Rates")
hist(DT[rate < 20]$rate, breaks = 50, main = "Distribution of Crash Rates Under 20")

hist(DT$mvmt, main = "Distribution of VMT") # VMT has an exponential distribution
hist(DT$annual_crashes, main = "Distribution of Crash Counts") # annual crashes has similar distribution
hist(log(DT$mvmt), main = "Distribution of Log VMT")
```


## Modeling

Having completed our exploratory data analysis, we see that there's a reliable relationship between crash counts and VMT. To estimate the lighting relationship, we'll fit a regression model controlling for VMT.

Count data is typically Poisson or Negative Binomially distributed, so we'll use an appropriate regression model.

```{r}
DT[, total_crashes := as.integer(annual_crashes * 5)] # total 5 year crashes for count model (whole numbers)

# fit a Poisson regression model
fit_p <- glm(total_crashes ~ lighting + mvmt, data = DT, family = poisson)
summary(fit_p)
# Dispersion Statistic
sum(resid(fit_p, type = "pearson")^2) / (nrow(DT) - length(coef(fit_p))) # data is 10x overdispersed
# goodness of fit measure:
1 - pchisq(summary(fit_p)$deviance, summary(fit_p)$df.residual) # model does not fit the data (p < 0.05)

# negative binomial regression model
fit_nb <- MASS::glm.nb(total_crashes ~ lighting + mvmt, data = DT)
summary(fit_nb)
# Dispersion Statistic
sum(resid(fit_nb)^2) / (nrow(DT) - length(coef(fit_nb)) + 1) # still overdispersed (1.13), but not as bad
# goodness of fit measure:
1 - pchisq(summary(fit_nb)$deviance, summary(fit_nb)$df.residual) # model does not fit the data

# zero inflated poisson model
fit_zip <- pscl::zeroinfl(total_crashes ~ lighting + mvmt | CUR_AADT, data = DT, dist = "poisson")
summary(fit_zip)
# Dispersion Statistic
sum(resid(fit_zip, type = "pearson")^2) / (nrow(DT) - length(coef(fit_zip))) # significant overdispersed (6x)
# goodness of fit:
1 - pchisq(sum(resid(fit_zip)^2), summary(fit_zip)$df.residual) # model does not fit the data (p < 0.05)

# zero inflated negative binomial model
fit_zinb <- pscl::zeroinfl(total_crashes ~ lighting + mvmt | CUR_AADT, data = DT, dist = "negbin")
summary(fit_zinb)
# Dispersion Statistic
sum(resid(fit_zinb, type = "pearson")^2) / (nrow(DT) - length(coef(fit_zinb)) + 1) # this is close to 1, which looks good
# goodness of fit:
1 - pchisq(sum(resid(fit_zinb)^2), summary(fit_zinb)$df.residual) # model doesn't fit but is closer than the others


# what does this look like?
synth <- data.table(lighting = c(rep("lit", 100), rep("unlit", 100)),
                    mvmt = seq(0.1, 70, length.out = 100))
fit_aadt <- lm(CUR_AADT ~ poly(mvmt, 3), data = DT)
synth$CUR_AADT <- predict(fit_aadt, newdata = synth)
synth$total_crashes <- predict(fit_zinb, newdata = synth)
ggplot(synth, aes(x = mvmt, y = total_crashes, color = lighting)) + 
  geom_point(data = DT, size = 0.5, alpha = 0.33) + 
  geom_line() +
  coord_cartesian(xlim = c(0, 40), ylim = c(0, 60)) + 
  ggtitle("Estimated Crashes and VMT", "By Lighting Condition")
       
```

# Conclusion

The zero-inflated negative binomial model outputs show that unlit road segments
have 0.47 fewer crashes on average than lit segments, controlling for VMT.

This finding is rather counter-intuitive from a "lighting improves safety"
perspective, but makes sense if lighting is installed in locations with higher
crash risk (e.g. intersections). 

More analysis and a richer dataset may be necessary to be able to isolate the safety
impact of lighting. For example, a dataset perhaps a maintenance dataset could be used
to determine when normally lit intersections were, in fact, dark—and compare the crash history
for identical locations under lit and unlit conditions.